Developing a Bubble Chamber Particle Discriminator Using Semi-Supervised Learning

### 2018-08-21 documentation/paper.tex
Minor edits throughout; added paragraph at end

\begin{abstract}
    The identification of non-signal events is a major hurdle to overcome for bubble chamber dark matter experiments such as PICO-60. The current practice of manually developing a discriminator function to eliminate background events is difficult when calibration data is frequently impure and present only in small quantities. In this study, several different input formats and neural network architectures are applied to the task. First, they are optimized in a supervised learning context. Next, novel semi-supervised learning algorithms are trained, and found to replicate the Acoustic Parameter discriminator previously used in PICO-60 with up to 100\% accuracy.
\end{abstract}

@@

    \item Determine whether these techniques can be extended, applying novel algorithms based on semi-supervised learning, to develop an effective discriminator function based on incomplete and inaccurate initial information.

    For clarity, and with specific reference to the PICO-60 experiment, the intent is to accurately separate particle types using only calibration data for training. While calibration runs are known to be mostly correlated with particle types, the relatively large percentage of impurities has the potential to hinder conventional learning techniques.
\end{enumerate}

@@

A significant challenge in this application is that neither supervised nor unsupervised learning is ideal. The calibration sets available for training are not pure, so supervised learning is likely to overfit on biases and produce undesirable results. Conversely, unsupervised learning may be able to find clusters in the data, but it is impossible to guarantee that it will distinguish between nuclear recoils and alpha particles as opposed to some other binary separation.

@@

Two semi-supervised algorithms were developed and implemented for this study: gravitational differentiation and iterative cluster nucleation. Both are discussed in detail in Section \ref{semi_supervised}.

@@

Classification accuracy is the ability of the network to separate the events into the two desired classes. Numerically, this corresponds to the number of validation examples the network classifies correctly.

@@

Numerically, this is a variable defined as $C$ below, where $N$ and $A$ are the sets of outputs of the supervised learning discriminator in question, corresponding to the sets of neutrons and alpha particles respectively. It calculates the spread of the network's predictions for each ground truth class. This means that a decisive discriminator, which produces a wide separation between the two classes, is preferred over one that produces a nebulous cloud of outputs with a seemingly arbitrary decision boundary.

@@

The first step is to calculate the standard deviation S of the union of the two sets. This gives an indication of the scale of the overall distribution. When $N$ and $A$ are divided by $S$, they are normalized so that the standard deviation of their union is equal to 1. While the neural network’s outputs are bounded in the range of 0 to 1 with a sigmoid activation, AP has a significantly wider range. Normalization of the union prevents this from creating a bias where AP would produce a higher standard deviation with a similarly proportioned error.

The second step is to calculate the mean of the standard deviations of the normalized sets of neutrons and alpha particles individually. This is an indication of how tightly clustered or widely dispersed the discriminator’s predictions are. Very consistent predictions of $x$ for neutrons and $y$ for alphas (for any $x$ and $y$), with minimal variance off those specific values, will produce a low class-wise standard deviation.

@@

All supervised learning configurations were trained and evaluated based on their ability to determine the origin of events: whether they are from calibration runs (predominantly neutrons) or background runs (predominantly alpha particles). This metric was used for training and testing instead of AP for two reasons:

@@

During early tests with variations on $DeepConv(\omega)$ without $WallCut(D)$, it was revealed to have a surprising and undesirable capacity to separate run types based on meaningless information. It was able to separate the AmBe calibration run from combined background runs with 96\% accuracy, and to separate all calibration runs from background runs with 91\% accuracy. This was determined to be caused by the network detecting and fitting on biases in the white noise at the beginning and hydraulic recompression sounds at the end of the raw waveform. This extraneous information was subsequently cropped out.

@@

In the context of PICO-60, the labeled sets consist of:

\begin{enumerate}
    \item Single-bubble events from the calibration sets, which are known to contain approximately 90\% neutrons
    \item Single-bubble events from the background radiation data sets, which contain approximately 99\% alpha particles
\end{enumerate}

In general, these labeled sets should have a strong but not perfect correlation with particle types. The unlabeled data can consist of any mixture of particle types.

@@

Iterative cluster nucleation is a second semi-supervised learning algorithm developed as part of this study. It takes advantage of some amount of imperfectly labeled data, using it to classify the rest of an unlabeled training set while optimizing to produce an effective discriminator.

@@

This study has demonstrated that it is possible, using semi-supervised learning, to automatically optimize a discriminator with performance comparable to one developed and tuned by humans like the Acoustic Parameter technique. This conclusion has been achieved through two rounds of optimization, the first to explore and evaluate various network structures and input formats using supervised learning, and the second to incorporate the most successful configuration into semi-supervised learning systems and replicate AP as accurately as possible.

@@

Two semi-supervised learning algorithms were developed, applying the best configuration found for supervised learning. While both techniques produced some very effective discriminators, iterative cluster nucleation was found to be the more effective of the two. It replicated AP with a mean of 98\% accuracy on the best configuration. This indicates that the technique shows significant promise, and may be worth consideration as a discrimination technique for future experiments.

\section{Software}

>>>>

\begin{abstract}
    The identification of non-signal events is a major hurdle to overcome for bubble chamber dark matter experiments such as PICO-60. The current practice of manually developing a discriminator function to eliminate background events is difficult when available calibration data is frequently impure and present only in small quantities. In this study, several different input formats and neural network architectures are applied to the task. First, they are optimized in a supervised learning context. Next, two novel semi-supervised learning algorithms are trained, and found to replicate the Acoustic Parameter discriminator previously used in PICO-60 with up to 100\% accuracy.
\end{abstract}

@@

    \item Determine whether these techniques can be extended, applying two original algorithms based on semi-supervised learning, to develop an effective discriminator function based on incomplete and inaccurate initial information.

    For clarity, and with specific reference to the PICO-60 experiment, the intent is to accurately separate particle types using only run types as training data. While run types are known to be mostly correlated with particle types, the relatively large percentage of impurities has the potential to hinder conventional learning techniques. Specifically, the run types available for training are:

    \begin{enumerate}
        \item Neutron calibration sets, which are known to contain approximately 90\% neutrons, and
        \item Background radiation runs, which contain approximately 99\% alpha particles
    \end{enumerate}
\end{enumerate}

@@

A significant challenge in this application is that neither supervised nor unsupervised learning is ideal. The calibration and background runs available for training are not pure, so supervised learning is likely to overfit on biases and produce undesirable results. Conversely, unsupervised learning may be able to find clusters in the data, but it is impossible to guarantee that it will distinguish between nuclear recoils and alpha particles as opposed to some other binary separation.

@@

Two original semi-supervised algorithms were developed and implemented for this study: gravitational differentiation and iterative cluster nucleation. Both are discussed in detail in Section \ref{semi_supervised}.

@@

Classification accuracy is the ability of the network to separate the events into the two desired classes. Numerically, this corresponds to the number of validation examples the network classifies correctly. The baseline for correctness is the run type during supervised learning experiments, and the AP prediction for semi-supervised experiments.

@@

Numerically, this is a variable defined as $C$ below, where $N$ and $A$ are the sets of outputs of the supervised learning discriminator in question, corresponding to the sets of neutrons and alpha particles respectively (or calibration and background sets respectively, depending on which is used as ground truth data). It calculates the spread of the network's predictions for each ground truth class. This means that a decisive discriminator, which produces a wide separation between the two classes, is preferred over one that produces a nebulous cloud of outputs with a seemingly arbitrary decision boundary.

@@

The first step is to calculate the standard deviation $S$ of the union of $N$ and $A$. This gives an indication of the scale of the overall distribution. When $N$ and $A$ are divided by $S$, they are normalized so that the standard deviation of their union is equal to 1. While the neural network’s outputs are bounded in the range of 0 to 1 with a sigmoid activation, AP has a significantly wider range. Normalization of the union prevents this from creating a bias where AP would produce a higher standard deviation with a similarly proportioned error.

The second step is to calculate the mean of the standard deviations of the normalized sets of neutrons and alpha particles individually. This is an indication of how tightly clustered or widely dispersed the discriminator’s predictions are for each class. Very consistent predictions of $x$ for neutrons and $y$ for alphas (for any $x$ and $y$), with minimal variance off those specific values, will produce a low class-wise standard deviation.

@@

All supervised learning configurations were trained and evaluated based on their ability to determine the origin of events: whether they are from neutron calibration source runs (predominantly neutrons) or background radiation runs (predominantly alpha particles). This metric was used for training and testing instead of AP for two reasons:

@@

During early tests with variations on $DeepConv(\omega)$ without $WallCut(D)$, it was revealed to have a surprising and undesirable capacity to separate run types based on meaningless information. It was able to discriminate between events from the AmBe neutron calibration runs and events from alpha background runs with 96\% accuracy, and to separate all combined neutron calibration runs from alpha background runs with 91\% accuracy. This was determined to be caused by the network detecting and fitting on biases in the white noise at the beginning and hydraulic recompression sounds at the end of the raw waveform. This extraneous information was subsequently cropped out.

@@

In the context of PICO-60, the labeled sets consist of the neutron calibration runs, and the background radiation runs (which consist predominantly of alpha particles). In general, these labeled sets should have a strong but not perfect correlation with particle types. The unlabeled data can consist of any mixture of particle types.

@@

Iterative cluster nucleation is a second original semi-supervised learning algorithm developed as part of this study. It takes advantage of some amount of imperfectly labeled data, using it to classify the rest of an unlabeled training set while optimizing to produce an effective discriminator.

@@

This study has demonstrated that it is possible, using semi-supervised learning, to automatically optimize a discriminator with performance comparable to one developed and tuned by humans like the Acoustic Parameter technique. This conclusion has been achieved through two rounds of optimization, the first to explore and evaluate various network structures and input formats using supervised learning, and the second to incorporate the most successful configuration into semi-supervised learning systems with the goal of replicating AP as accurately as possible.

@@

Two semi-supervised learning algorithms were developed, applying the best configuration found for supervised learning. While both techniques produced some very effective discriminators, iterative cluster nucleation was found to be the more effective of the two. It replicated AP with a mean of 98\% accuracy on the best configuration. This indicates that the technique shows significant promise.

While there is still of course a place for conventionally developed discriminators such as AP, they are difficult to develop until a large quantity of calibration data has been collected and the physical principles of the experiment are understood well. The innovations developed in this study suggest that semi-supervised learning can facilitate very quick development of discriminators immediately after the first calibration runs are conducted on a new experimental apparatus. This has the potential to allow the teams working on these experiments to iterate more quickly and better understand intermediate results, before all data has been analyzed in depth.

\section{Technologies}
